---
title: "STAT577 Homework 2"
author: "Tim Farkas"
date: "3/19/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Packages
```{r Packages, message=FALSE}
library(tidyverse)
library(rjags)
library(R2jags)
library(kableExtra)
```

### Problem 1

##### Create Gamma-Buster
```{r gammabustr}

gammabustr <- function(mode, pct=0.95, target, 
                       beta_upper=100, plot=FALSE) {
  
  #### helper functions ####
  
  # function to find alpha given beta and mode
  find_alpha <- function(beta, mode) mode * beta + 1
  
  # function to get difference from boundary target
  diff_pct <- function(beta, mode, prop, target) {
    
      # function to get theta given percentile
      get_pct <- function(beta, mode, prop) {
        
        alpha <- find_alpha(beta=beta, mode=mode)
        percentile <- qgamma(prop, rate=beta, shape=alpha)
        return(percentile)
    
      }
    
    abs(get_pct(beta=beta, mode=mode, prop=prop) - target)
    
  }
  
  #### main function body ####
  par <- optim(par=1, fn=diff_pct, method="Brent", mode=mode, 
               prop=pct, target=target, lower=0, upper=beta_upper)$par
  
  parms <- c(alpha=find_alpha(beta=par, mode=mode), beta=par)
  
  #### plot ####
  
  # plot theoretical distribution given parameters
  if(plot) {
    
    gammas <- seq(qgamma(.0001, parms["alpha"], parms["beta"]),
                  qgamma(.9999, parms["alpha"], parms["beta"]),
                  length.out=10000)
    
    sims <- tibble(sims=dgamma(gammas, parms["alpha"], parms["beta"]))
    
    print(ggplot(data=sims, aes(x=gammas, y=sims)) + 
            geom_line() + 
            xlab("theta"))
    
  }
  
  #### return ####
  return(parms)
  
}

```

1a. A prior given Dr. McMillan's beliefs would be Gamma(2.31, 2.63). 
```{r}
gammabustr(.5, target=2, plot=TRUE)
```

1b. The conjugate posterior is Gamma($\sum{x_i} + \alpha$, $n + \beta$). With 38 hunters catching 10 armadillos, that yields a posterior for the kill rate of Gamma(12.31, 40.63).

1c. The expected value of a Gamma(12.31, 40.63) is $\frac{\alpha}{\beta} = \frac{12.31}{40.63} = 0.303$.

1d. There is a 95% probability that the kill rate lies between 0.156 and 0.493 armadillos per day. With 38 hunters, that equates to an estimate of between 6 and 18 armadillos.

```{r}
sims <- rgamma(100000, 12.06, 40.12)
up_low <- quantile(sims, c(0.025, 0.975)) 
up_low * 38
```  

### Bullet 2  

```{r JAGS model, error=TRUE, eval=FALSE}
sink("~/Dropbox/stat577/homework/hw_code/example_3.1.3.txt")

cat("
model{
      
  # likelihood distributions
  y1 ~ dbin(t1, 80)
  y2 ~ dbin(t2, 100)
  
  # prior distributions
  t1 ~ dbeta(1, 1)
  t2 ~ dbeta(2, 1)
  
}", fill=TRUE)

sink()
```

```{r run model, message=FALSE, results=FALSE}
params <- c("t1", "t2")

inits <- function() {
  list(
    t1 = rbeta(1, 1, 1), 
    t2 = rbeta(1, 2, 1)
  )
}

jags_data <- list(y1 = 32, 
                  y2 = 35)

model <- "~/Dropbox/stat577/homework/hw_code/example_3.1.3.txt"

fit <- jags(jags_data, inits, params, model)

```
### Problem 2

$$y_i \sim Poisson(\theta M_i)$$
$$\theta \sim Gamma(\alpha, \beta)$$
$$p(y | \theta) = \prod_{i = 1}^{n} \left[ \frac{\theta^{y_i}e^{-\theta M_i}}{y_i!}\right] = \frac{e^{-\theta \sum{M_i}}\theta^{\sum{y_i}}}{\prod{y_i!}}$$
$$= \frac{\left[\frac{e^{-\theta \sum{M_i}}\theta^{\sum{y_i}}}{\prod{y_i!}}\right]\left[\frac{\beta^{\alpha}}{\Gamma(\alpha)}\theta^{\alpha-1}e^{-\beta\theta}\right]}{\int_0^\infty {\left[\frac{e^{-\theta \sum{M_i}}\theta^{\sum{y_i}}}{\prod{y_i!}}\right]\left[\frac{\beta^{\alpha}}{\Gamma(\alpha)}\theta^{\alpha-1}e^{-\beta\theta}\right]}d\theta}$$
$$= \frac{\left[e^{-\theta \sum{M_i}}\theta^{\sum{y_i}}\right]\left[\theta^{\alpha-1}e^{-\beta\theta}\right]}{\int_0^\infty {\left[e^{-\theta \sum{M_i}}\theta^{\sum{y_i}}\right]\left[\theta^{\alpha-1}e^{-\beta\theta}\right]}d\theta}$$
$$= \frac{e^{-\theta(\beta + \sum{M_i})}\theta^{(\alpha+\sum{y_i}) - 1}}{\int_0^\infty e^{-\theta(\beta + \sum{M_i})}\theta^{(\alpha+\sum{y_i}) - 1}d\theta}$$
$$= \frac{e^{-\theta(\beta + \sum{M_i})}\theta^{(\alpha+\sum{y_i}) - 1}}{\frac{\Gamma(\alpha + \sum{y_i})}{(\beta + \sum{M_i})^{\alpha + \sum{y_i}}}\int_0^\infty \frac{(\beta + \sum{M_i})^{\alpha + \sum{y_i}}}{\Gamma(\alpha + \sum{y_i})}e^{-\theta(\beta + \sum{M_i})}\theta^{(\alpha+\sum{y_i}) - 1}d\theta}$$
$$= \frac{(\beta + \sum{M_i})^{\alpha + \sum{y_i}}}{\Gamma(\alpha + \sum{y_i})}e^{-\theta(\beta + \sum{M_i})}\theta^{(\alpha+\sum{y_i}) - 1}$$
$$\therefore p(\theta|y) \sim Gamma(\alpha + \sum{y_i}, \beta + \sum{M_i}) $$
  
  
### Problem 3  

$$y_i \sim Poisson(\theta_i M_i)$$
$$\theta_i \sim Gamma(\alpha_i, \beta_i)$$
$$p(\theta_1, \theta_2) = \left[\frac{\beta_1^{\alpha_1}}{\Gamma(\alpha_1)}\right]
\theta_1^{y_1}
e^{-\beta_1\theta_1}
\left[\frac{\beta_2^{\alpha_2}}{\Gamma(\alpha_2)}\right]
\theta_2^{y_2}
e^{-\beta_2\theta_2} $$

$$p(y_1, y_2|\theta_1, \theta_2) = 
\frac{1}{y_1!}
\theta_1^{y_1}
e^{-\theta_1M_1}
\frac{1}{y_2!}
\theta_2^{y_2}
e^{-\theta_2M_2}$$  


$$p(\theta_1, \theta_2|y_1, y_2) = 
\frac{
\left[\frac{\beta_1^{\alpha_1}\beta_2^{\alpha_2}}{\Gamma(\alpha_1)\Gamma(\alpha_2)}\right]
\left[\frac{1}{y_1!y_2!}\right]
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
}
{
\left[\frac{\beta_1^{\alpha_1}\beta_2^{\alpha_2}}{\Gamma(\alpha_1)\Gamma(\alpha_2)}\right]
\left[\frac{1}{y_1!y_2!}\right]
\int_0^{\infty}
\int_0^{\infty}
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
d\theta_1 d\theta_2
}$$

$$= 
\frac{
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
}
{
\int_0^{\infty}
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
d\theta_1
\int_0^{\infty}
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
d\theta_2
}$$

$$=\frac{
\left[
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
\right]
\left[
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
\right]}
{\left[\frac{\Gamma(\alpha_1 + y_1)}{(\beta_1 + M_1)^{\alpha_1 + y_1}}\right]
\left[
\int_0^{\infty}
\left(\frac{[\beta_1 + M_1]^{\alpha_1 + y_1}}{\Gamma[\alpha_1 + y_1]}\right)
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
d\theta_1
\right]
\left[\frac{\Gamma(\alpha_2 + y_2)}{(\beta_2 + M_2)^{\alpha_2 + y_2}}\right]
\left[
\int_0^{\infty}
\left(\frac{[\beta_2 + M_2]^{\alpha_2 + y_2}}{\Gamma[\alpha_2 + y_2]}\right)
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
d\theta_2
\right]}$$

$$ = 
\frac{
\left[
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
\right]
\left[
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
\right]
}
{
\left[\frac{\Gamma(\alpha_1 + y_1)}{(\beta_1 + M_1)^{\alpha_1 + y_1}}\right]
\left[\frac{\Gamma(\alpha_2 + y_2)}{(\beta_2 + M_2)^{\alpha_2 + y_2}}\right]
}$$


$$= 
\left[\frac{(\beta_1 + M_1)^{\alpha_1 + y_1}}{\Gamma(\alpha_1 + y_1)}\right]
\left[
\theta_1^{\alpha_1 + y_1 - 1}
e^{-\theta_1(\beta_1 + M_1)}
\right]
\left[\frac{(\beta_2 + M_2)^{\alpha_2 + y_2}}{\Gamma(\alpha_2 + y_2)}\right]
\left[
\theta_2^{\alpha_2 + y_2 - 1}
e^{-\theta_2(\beta_2 + M_2)}
\right]$$  

$$\therefore 
p(\theta_1, \theta_2|y_1, y_2) \sim 
Gamma(\alpha_1 + y_1, \beta_1 + M_1) 
\times
Gamma(\alpha_2 + y_2, \beta_2 + M_2) $$

$$\sim 
Gamma(501, 100.1)
\times
Gamma(801, 200.1)$$  


### Problem 4  

```{r, eval=FALSE, error=TRUE}
sink("~/Dropbox/stat577/homework/hw_code/problem4.txt")

cat("
model{
    
  for(i in 1:n) {
  
    # likelihoods
    y[i] ~ dpois(t[i] * M[i])
    
    # priors
    t[i] ~ dgamma(1, 0.1)
    
  }
  
}", fill=TRUE)

sink()
```

```{r, message=FALSE, results=FALSE}
params <- c("t")

inits <- function() {
  list(
    t = rgamma(2, 1, 0.1)
  )
}

jags_data <- list(y = c(500, 800), 
                  M = c(100, 200),
                  n = 2)

model <- "~/Dropbox/stat577/homework/hw_code/problem4.txt"

fit.4 <- jags(jags_data, inits, params, model,
            n.iter=10000, n.burnin=1000)
```

```{r}

apply(fit.4$BUGSoutput$sims.list$t, 2, mean)
```

The means of theoretical Gamma distributions with alphas and betas of 500 & 800 and 100 and 200, respectively, are $\frac{\alpha}{\beta} = \frac{501}{100.1}$ and $\frac{801}{200.1}$ = 5.005 and 4.002. The mean of posteriors from JAGS output are 5.007 ad 4.003, respectively. Very close. 

### Problem 5

##### With Jeffrey's Prior: Beta(0.5, 0.5)

```{r, eval=FALSE, error=TRUE}
sink("~/Dropbox/stat577/homework/hw_code/hw2_problem5.1.txt")

cat("
model{
  
    # likelihoods
    y[1] ~ dbin(theta1, n[1])
    y[2] ~ dbin(theta2, n[2])
  
    # priors
    theta2 ~ dbeta(.5, .5)
    delta ~ dnorm(0, .5)
    
    # derived quantities
    theta1 <- exp(delta) * theta2 / (1 - theta2 * (1 - exp(delta)))
    OR <- theta1 / (1 - theta1) / (theta2 / (1 - theta2))
  
  
}", fill=TRUE)

sink()
```

##### With Beta(1, 1) Prior 

```{r, eval=FALSE, error=TRUE}
sink("~/Dropbox/stat577/homework/hw_code/hw2_problem5.2.txt")

cat("
model{
    

  
    # likelihoods
    y[1] ~ dbin(theta1, n[1])
    y[2] ~ dbin(theta2, n[2])
  
    # priors
    theta2 ~ dbeta(1, 1)
    delta ~ dnorm(0, .5)
    
    # derived quantities
    theta1 <- exp(delta) * theta2 / (1 - theta2 * (1 - exp(delta)))
    OR <- theta1 / (1 - theta1) / (theta2 / (1 - theta2))
  
  
}", fill=TRUE)

sink()
```

##### With Uniform(log(0.02), log(50))  Prior
```{r, eval=FALSE, error=TRUE}
sink("~/Dropbox/stat577/homework/hw_code/hw2_problem5.3.txt")

cat("
model{
  
    # likelihoods
    y[1] ~ dbin(theta1, n[1])
    y[2] ~ dbin(theta2, n[2])
  
    # priors
    theta2 ~ dunif(log(.02), log(50))
    delta ~ dnorm(0, .5)
    
    # derived quantities
    theta1 <- exp(delta) * theta2 / (1 - theta2 * (1 - exp(delta)))
    OR <- theta1 / (1 - theta1) / (theta2 / (1 - theta2))
  
  
}", fill=TRUE)

sink()
```

```{r, message=FALSE, results=FALSE}
params <- c("theta1",
            "theta2", 
            "OR")

inits <- function() {
  list(
    theta2 = rbeta(1, 1, 1),
    delta = rnorm(1, 0, 2)
  )
}

jags_data <- list(y = c(7, 8), 
                  n = c(7, 16))

model1 <- "~/Dropbox/stat577/homework/hw_code/hw2_problem5.1.txt"
model2 <- "~/Dropbox/stat577/homework/hw_code/hw2_problem5.2.txt"
model3 <- "~/Dropbox/stat577/homework/hw_code/hw2_problem5.3.txt"

fit1 <- jags(jags_data, 
            inits,
            params, model1,
            n.iter=10000, n.burnin=1000)

fit2 <- jags(jags_data, 
            inits,
            params, model2,
            n.iter=10000, n.burnin=1000)

fit3 <- jags(jags_data, 
            inits,
            params, model3,
            n.iter=10000, n.burnin=1000)

```

```{r, echo=FALSE}
outmat <- matrix(
  c(
    mean(fit1$BUGSoutput$sims.list$OR > 1),
    mean(fit2$BUGSoutput$sims.list$OR > 1),
    mean(fit3$BUGSoutput$sims.list$OR > 1),
    mean(fit1$BUGSoutput$sims.list$OR > 2),
    mean(fit2$BUGSoutput$sims.list$OR > 2),
    mean(fit3$BUGSoutput$sims.list$OR > 2)), 
  nrow=3, dimnames=list(c("Jeffrey's", "Beta(1, 1)",
                          "U(ln(0.02), ln(50))")))
kable_styling(kable(round(outmat, 3) ,
col.names = c("p(OR > 1)", "p(OR > 2)")))
```


The results are not particularly sensitive to these choices of prior. The probability that the odds ratio is greater than 1 is between 0.983 and 0.985, and the probability that the odds ratio is greater than 2 is between .89 and .91.



